---
title: "Linear Models Project in R"
subtitle: "Predicting Student Academic Performance"
author:
  - name: "BEN SAID Daniel ; DESCHILDRE Mateo ; ROY Damien"
    affiliation: "M1–MIDO, Université Paris Dauphine-PSL"
date: last-modified
lang: en
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    toc-title: "Contents"
    number-sections: true
    smooth-scroll: true
    page-layout: full
    linestretch: 1.2
    fontsize: 125%
    self-contained: true
    df-print: default
    code-fold: true
    code-tools: true
    code-line-numbers: false
    code-overflow: scroll
    code-copy: true
    highlight-style: github
    theme: cosmo
    code-block-bg: "#F5F5F5"
    html-math-method: mathjax
    fig-format: retina
    fig-width: 13
    fig-height: 7
    fig-align: center
    fig-pos: "H"
    fig-cap-location: bottom
    mainfont: "Source Sans Pro"
execute:
  echo: true
  warning: false
  message: false
  error: false
editor:
  markdown:
    wrap: 110
    canonical: true
knitr:
  opts_chunk:
    tidy: styler
    comment: ''
    dev: svg
---

```{r}
#| label: setup
#| include: false

# Reproducibility
set.seed(42)

# Global options
options(pillar.width = 130, width = 200, scipen = 999, digits = 5)

# Load packages
library("broom")
library("car")
library("caret")
library("collapse")
library("correlation")
library("datawizard")
library("effectsize")
library("GGally")
library("ggfortify")
library("ggpubr")
library("ggrepel")
library("glue")
library("gtsummary")
library("here")
library("insight")
library("janitor")
library("kableExtra")
library("lmtest")
library("marginaleffects")
library("matrixTests")
library("modelbased")
library("multcomp")
library("openxlsx")
library("parameters")
library("patchwork")
library("performance")
library("pROC")
library("qqplotr")
library("rstatix")
library("scales")
library("see")
library("tidyverse")
library("interactions")

source(here("script", "helper_functions.R"))
```

# Introduction

## Context and Research Problem

Academic performance is influenced by a complex interplay of factors including study habits, living
conditions, socio-educational context, and health-related behaviors. Understanding these relationships is
crucial for educators, policymakers, and students themselves to identify actionable interventions that can
improve educational outcomes.

This project aims to predict student academic performance (measured by exam scores) using multiple linear
regression, while exploring the relative importance of various predictors and potential interactions between
them. The analysis follows a dual objective: building an **interpretable model** for inference and
understanding relationships, and achieving **strong predictive performance** on unseen data.

## Dataset Overview

The dataset comprises **5,000 observations** collected from students, with **15 variables** covering different
dimensions of their academic journey and daily life:

**Life habits and well-being:**

-   `sleep_hrs` (continuous): nightly sleep duration
-   `sleep_qual` (ordinal): subjective sleep quality (Poor, Average, Good)
-   `trav_time` (ordinal): daily commute time to school

**Study habits:**

-   `study_hrs` (continuous): weekly study hours
-   `study_method` (categorical): preferred study method (6 categories)
-   `attend_pct` (continuous): percentage of class attendance

**Contextual and environmental factors:**

-   `age` (continuous): student age in years
-   `sexe` (categorical): gender (Female, Male, Other)
-   `school_type` (categorical): type of institution (Public, Private)
-   `web_access` (categorical): home internet access (Yes, No)
-   `parent_educ` (ordinal): parental education level (6 levels)
-   `extra_act` (categorical): participation in extracurricular activities (Yes, No)

The dependent variable `y` represents a quantitative measure of academic performance (exam score, range
0-100).

## Research Hypotheses

Based on existing literature in education science and cognitive psychology, we formulate the following
hypotheses:

**H1: Non-linear effect of sleep**\
Sleep duration (`sleep_hrs`) exhibits a quadratic (inverted-U) relationship with performance, where both
insufficient and excessive sleep negatively affect scores.

**H2: Interaction between study quantity and sleep quality**\
The effectiveness of study time (`study_hrs`) depends on sleep quality (`sleep_qual`). Good sleep should
amplify the benefits of studying.

**H3: Interaction between family background and school type**\
The effect of parental education (`parent_educ`) on performance varies by school type (`school_type`),
reflecting different socio-educational contexts.

**H4: Main effects**\
- Higher attendance (`attend_pct`) should positively predict performance - Internet access (`web_access`)
facilitates access to educational resources - Extracurricular activities (`extra_act`) may develop
transferable skills - Excessive commute time (`trav_time`) negatively impacts available study and rest time

# Methods

## Data Management and Preparation

### Data Import and Structure

```{r}
#| label: data-import
#| results: hide

data <- read.csv(here("data", "project.csv"))

cat("Sample size:", nrow(data), "observations\n")
cat("Number of variables:", ncol(data), "\n\n")

glimpse(data)
```

-   **Sample size**: 5000 observations

-   **Number of variables**: 16

### Variable Conversion and Labeling

We convert coded categorical variables into properly labeled factors with meaningful reference categories:

```{r}
#| label: data-cleaning

data_clean <- data |>
  mutate(
    sexe = factor(sexe, levels = c(1, 2, 3), labels = c("Female", "Male", "Other")),
    school_type = factor(school_type, levels = c(1, 2), labels = c("Public", "Private")),
    parent_educ = factor(parent_educ, levels = 1:6, 
                         labels = c("No formal", "High school", "Graduate", 
                                    "Postgraduate 1", "Postgraduate 2", "PhD"), 
                         ordered = TRUE),
    sleep_qual = factor(sleep_qual, levels = c(1, 2, 3), 
                        labels = c("Poor", "Average", "Good"), ordered = TRUE),
    web_access = factor(web_access, levels = c(1, 2), labels = c("No", "Yes")),
    trav_time = factor(trav_time, levels = 1:4, 
                       labels = c("<15 min", "15–30 min", "30–60 min", ">60 min"), 
                       ordered = TRUE),
    extra_act = factor(extra_act, levels = c(1, 2), labels = c("No", "Yes")),
    study_method = factor(study_method, levels = 1:6, 
                          labels = c("Online videos", "Coaching", "Notes", 
                                     "Textbook", "Group study", "Mixed"))
  ) |>
  # Set reference categories
  mutate(
    sexe = relevel(sexe, ref = "Female"),
    school_type = relevel(school_type, ref = "Public"),
    web_access = relevel(web_access, ref = "No"),
    extra_act = relevel(extra_act, ref = "No"),
    study_method = relevel(study_method, ref = "Online videos")
  ) |>
  # Apply descriptive labels
  relabel(
    sexe = "Gender",
    school_type = "School type",
    parent_educ = "Parental education",
    sleep_qual = "Sleep quality",
    web_access = "Internet access",
    trav_time = "Commute time",
    extra_act = "Extra curricular activities",
    study_method = "Study method",
    y = "Exam score",
    age = "Age (years)",
    attend_pct = "School attendance (%)"
  )
```

**Reference categories:**

-   `school_type`: Public (baseline)
-   `web_access`: No (baseline)
-   `extra_act`: No (baseline)
-   `study_method`: Online videos (baseline)
-   Ordinal variables maintain their natural ordering

### Vizualisation of quantitative variables

```{r}
#| label: data-viz

numeric_vars <- c("y", "age", "study_hrs", "sleep_hrs", "attend_pct")
vlabels <- c("Exam score", "Age (years)", "Weekly study (hours)", 
             "Sleep duration (hours)", "School attendance (%)")

data_clean |>
  select(id, all_of(numeric_vars)) |>
  pivot_longer(cols = -id, names_to = "var", values_to = "value") |>
  mutate(var = factor(var, levels = numeric_vars, labels = vlabels)) |>
  ggplot(aes(x = value)) +
  geom_histogram(fill = "dodgerblue", color = "black", bins = 20, linewidth = 0.5) +
  facet_wrap(~var, scales = "free") +
  scale_y_continuous(expand = expansion(c(0, 0.05))) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(x = "Value", y = "Count") +
  theme_bw(base_size = 14) +
  theme(strip.text = element_text(size = 11, face = "bold"))
```

### Data Integrity and Plausibility Checks

**Duplicate check**

```{r}
#| label: data-integrity1

# Check for duplicates
duplicates <- data_clean |> filter(duplicated(id))
cat("Number of duplicate IDs:", nrow(duplicates), "\n\n")
```

**Missing values**

```{r}
#| label: data-integrity2

# Check for missing values
na_count <- colSums(is.na(data_clean))
if(sum(na_count) == 0) {
  cat("✓ No missing values detected\n\n")
} else {
  print(na_count[na_count > 0])
}
```

**Range verification**

```{r}
#| label: data-integrity3


# Range verification
data_clean |>
  summarise(
    across(c(y, age, study_hrs, sleep_hrs, attend_pct),
           list(min = min, max = max, mean = mean, sd = sd))
  ) |>
  pivot_longer(everything()) |>
  separate(name, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") |>
  pivot_wider(names_from = Statistic, values_from = value) |>
  kable(digits = 2, caption = "Descriptive statistics for numeric variables")
```

**Implausible values**

```{r}
#| label: data-integrity4


# Check for implausible values
invalid_data <- data_clean |>
  filter(
    y < 0 | y > 100 |
    sleep_hrs < 0 | sleep_hrs > 24 |
    study_hrs < 0 | study_hrs > 168 |
    attend_pct < 0 | attend_pct > 100 |
      
    age < 0 |
    sexe < 1 | sexe > 3 |
    school_type < 1 | school_type > 2 |
    parent_educ < 1 | parent_educ > 6 |
    sleep_qual < 1 | sleep_qual > 3 |
    web_access < 1 | web_access > 2 |
    trav_time < 1 | trav_time > 4 |
    extra_act < 1 | extra_act > 2 |
    study_method < 1 | study_method > 6 |
    attend_pct_cat < 1 | attend_pct_cat > 4 
  )
cat("Number of implausible values:", nrow(invalid_data), "\n")
```

**Findings:** The dataset shows no missing values, no duplicate IDs, and all values fall within plausible
ranges. This indicates high data quality suitable for modeling.

### Handling Redundant Representations

```{r}
#| label: redundancy-analysis
#| fig-cap: "Comparison of continuous vs categorical versions of age and attendance"

p1 <- ggplot(data_clean, aes(x = factor(agecat), y = age)) +
  geom_boxplot(fill = "grey80") +
  labs(title = "Age vs Age category", x = "Age category", y = "Age (years)") +
  theme_bw()

p2 <- ggplot(data_clean, aes(x = factor(attend_pct_cat), y = attend_pct)) +
  geom_boxplot(fill = "grey80") +
  labs(title = "Attendance vs Attendance category",
       x = "Attendance category", y = "Attendance (%)") +
  theme_bw()

p1 + p2

# Quantify redundancy
age_cor <- cor(data_clean$age, as.numeric(data_clean$agecat), method = "spearman")
attend_cor <- cor(data_clean$attend_pct, as.numeric(data_clean$attend_pct_cat), 
                  method = "spearman")

cat("Spearman correlation - Age vs AgeCat:", round(age_cor, 3), "\n")
cat("Spearman correlation - Attend vs AttendCat:", round(attend_cor, 3), "\n")
```

**Decision:** Both pairs show near-perfect correlation (r \> 0.99), indicating they represent identical
information. We **retain the continuous versions** (`age`, `attend_pct`) to preserve granular information and
avoid multicollinearity. The categorical versions (`agecat`, `attend_pct_cat`) are removed from the analysis.

## Modeling Strategy

Our modeling approach follows a systematic progression:

1.  **Exploratory Data Analysis (EDA):** Understand distributions, relationships, and patterns
2.  **Baseline models:** Simple regressions to establish initial associations
3.  **Candidate model development:** Build multiple specifications with justified predictors
4.  **Model selection:** Compare using nested F-tests, AIC/BIC, and adjusted R²
5.  **Diagnostics:** Verify assumptions and check for influential observations
6.  **Final model:** Select parsimonious, interpretable model with strong predictive performance

**Validation protocol:** 80/20 train-test split with `set.seed(42)` for reproducibility. All preprocessing
decisions are made on training data only to avoid data leakage.

# Results

## Exploratory Data Analysis

### Outcome Distribution

```{r}
#| label: outcome-distribution
#| fig-cap: "Distribution of exam scores showing approximately symmetric distribution"

p_hist <- ggplot(data_clean, aes(x = y)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "dodgerblue", color = "white") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Histogram and Density of Exam Score", x = "Score (y)", y = "Density") +
  theme_minimal()

p_box <- ggplot(data_clean, aes(y = y)) +
  geom_boxplot(fill = "dodgerblue", outlier.color = "red") +
  labs(title = "Boxplot of Exam Score", y = "Score (y)") +
  theme_minimal()

p_hist + p_box

skew_val <- skewness(data_clean$y)$Skewness

n_outliers <- sum(
  abs(scale(data_clean$y)) > 3
)

cat("Skewness:", round(skew_val, 3), "\n")
cat("Interpretation:", 
    ifelse(abs(skew_val) < 0.5, "approximately symmetric", 
           ifelse(skew_val > 0, "right-skewed", "left-skewed")), "\n")
cat("Outliers:", n_outliers, "detected (within plausible [0, 100] range)\n")
```

**Key findings:**

-   Exam scores are approximately normally distributed (skewness ≈ 0)
-   Range: 0-100 with few outliers
-   No transformation needed for OLS regression

### Descriptive Statistics

```{r}
#| label: descriptive-stats

data_clean |>
  select(y, age, study_hrs, sleep_hrs, attend_pct) |>
  describe_distribution(centrality = c("mean", "median"), quartiles = TRUE) |>
  as_tibble() |>
  kable(digits = 2, caption = "Descriptive statistics for quantitative variables")

tab_freq1(data_clean, c("sexe", "school_type", "parent_educ", "sleep_qual", "web_access",
                 "trav_time", "extra_act", "study_method"), digits= 1) |>
  kable(align = "l", padding= 2, caption = "Frequency tables for categorial variables")
```

### Predictor-Outcome Relationships

#### Quantitative Predictors

```{r}
#| label: quant-predictors
#| fig-cap: "Scatterplots with LOESS smooths to assess linearity. Strong positive trends for study_hrs and attend_pct; moderate for sleep_hrs; weak for age."
#| fig-height: 8

data_clean |>
  select(y, age, study_hrs, sleep_hrs, attend_pct) |>
  pivot_longer(-y, names_to = "predictor", values_to = "value") |>
  ggplot(aes(x = value, y = y)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  facet_wrap(~predictor, scales = "free_x") +
  labs(title = "Exam Score vs Quantitative Predictors",
       subtitle = "Red line shows LOESS smooth to assess linearity",
       x = "Predictor value", y = "Exam Score") +
  theme_bw()
```

**Linearity assessment:**

-   **attend_pct:** Strong positive linear trend (excellent predictor)
-   **study_hrs:** Strong positive linear trend (excellent predictor)
-   **sleep_hrs:** Moderate positive trend with slight plateau at higher values
-   **age:** Nearly flat, highly scattered (very weak predictor, consider excluding)

#### Categorical Predictors

```{r}
#| label: cat-predictors
#| fig-cap: "Boxplots showing exam score distribution across categorical predictor levels. Red diamonds indicate group means."
#| fig-height: 10

cat_preds <- c("sexe", "school_type", "sleep_qual", "extra_act", "study_method")

data_clean |>
  select(y, all_of(cat_preds)) |>
  # Ensure all predictors are treated as characters for pivoting
  mutate(across(all_of(cat_preds), as.character)) |>
  pivot_longer(all_of(cat_preds), names_to = "var", values_to = "value") |>
  # Apply labels using the same logic as your Pok code
  mutate(var = factor(var, levels = cat_preds, labels = vlabels(data_clean[, cat_preds]))) |>
  # Shorten long labels (like study methods) if necessary
  mutate(value = fct_relabel(factor(value), \(x) str_trunc(x, 15))) |>
  ggplot(aes(x = y, y = value)) +
  # Boxplot with thinner lines for a cleaner look
  geom_boxplot(linewidth = 0.3, median.linewidth = 0.8, fill = "lightblue") +
  # Add mean point to anticipate regression coefficients
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "firebrick") +
  facet_wrap(vars(var), scales = "free_y") +
  labs(
    x = "Exam Score (y)", 
    y = NULL,
    title = "Bivariate Intuition: Categorical Predictors vs Score",
    subtitle = "Red diamonds represent the group mean (target of linear regression)"
  ) +
  theme_bw(base_size = 14) +
  labs_pubr() +
  theme(
    strip.text = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 9, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Key patterns:**

-   **web_access:** Large difference (\~7 points: No ≈ 50, Yes ≈ 57)
-   **extra_act:** Large difference (\~9 points: No ≈ 51, Yes ≈ 60)
-   **school_type:** Moderate difference (\~6 points: Public ≈ 53, Private ≈ 59)
-   **parent_educ:** Positive monotone trend (\~13 point range)
-   **trav_time:** Negative monotone trend (\~16 point range)
-   **sexe:** Minimal differences (all ≈ 56)

### Assessing Ordinal Predictors

```{r}
#| label: ordinal-assessment
#| fig-cap: "Monotonicity assessment for ordered categorical predictors. Lines connect group means with 95% confidence intervals."
#| fig-height: 8

p1 <- plot_monotone(data_clean, "sleep_qual", "Sleep Quality")
p2 <- plot_monotone(data_clean, "parent_educ", "Parental Education")
p3 <- plot_monotone(data_clean, "trav_time", "Commute Time")

(p1 | p2) / p3 +
  plot_annotation(
    title = "Assessing Monotonicity for Ordered Categorical Predictors",
    subtitle = "Points show mean ± 1.96×SE; red line shows overall trend"
  )
```

**Ordinal modeling decisions:**

-   **trav_time:** ✓ Strong monotone negative effect (use as ordered factor)
-   **parent_educ:** ✓ Clear positive monotone pattern with a potential plateau (use as ordered factor)
-   **sleep_qual:** ⚠ Positive monotone with a potential plateau (use as ordered with caution)

### Predictor Associations

We check for potential multicollinearity issues among predictors.

```{r}
#| label: multicollinearity-check
#| fig-cap: "VIF (Variance Inflation Factor) to detect multicollinearity. Values < 5 indicate no concern."

# 1. Fit the additive model
vif_model <- lm(y ~ age + sexe + school_type + web_access + study_hrs + 
                attend_pct + sleep_hrs + sleep_qual + trav_time + 
                parent_educ + extra_act + study_method, 
                data = data_clean)

# 2. Calculate VIF/GVIF
vif_values <- car::vif(vif_model)

# 3. Process and display results
if(is.matrix(vif_values)) {
  # Transformation explicite de la matrice en data.frame
  vif_df <- as.data.frame(vif_values)
  vif_df$Variable <- rownames(vif_df)
  
  # Nettoyage des noms de colonnes pour éviter les erreurs avec les caractères spéciaux
  colnames(vif_df) <- c("GVIF", "Df", "Adj_GVIF", "Variable")
  
  # Affichage
  print(vif_df %>%
    select(Variable, GVIF, Df, Adj_GVIF) %>%
    arrange(desc(Adj_GVIF)) %>%
    kable(digits = 2, caption = "Generalized VIF (GVIF) for predictors"))
    
  # Récupération du max pour la conclusion
  max_val <- max(vif_df$Adj_GVIF)
  
} else {
  # Cas des variables continues uniquement
  vif_df_simple <- data.frame(
    Variable = names(vif_values),
    VIF = vif_values
  )
  
  print(vif_df_simple %>%
    arrange(desc(VIF)) %>%
    kable(digits = 2, caption = "VIF values (threshold: 5-10)"))
    
  max_val <- max(vif_values)
}

# 4. Conclusion dynamique
cat("\n--- CONCLUSION ---\n")
if(max_val < 5) {
  cat(paste("✓ No multicollinearity concerns (Max Adjusted VIF =", round(max_val, 2), ")\n"))
} else {
  cat(paste("⚠ Warning: Some variables show moderate to high multicollinearity (Max =", round(max_val, 2), ")\n"))
}
```

**Finding:** No severe multicollinearity detected among predictors. All VIF values are well below the
threshold of 5, supporting their independent inclusion in models.

### EDA Conclusions

**Strong predictor candidates:**

1.  `attend_pct` (r ≈ 0.65): Strong linear correlation
2.  `study_hrs` (r ≈ 0.60): Strong linear correlation
3.  `web_access`: \~10 point difference
4.  `extra_act`: \~8 point difference

**Moderate candidates:**

-   `school_type`, `parent_educ`, `sleep_hrs`

**Weak/exclude:**

-   `age` (r ≈ 0.03): Very weak, consider excluding
-   `sexe`: No visible differences

**Interaction hypotheses to test:**

1.  `study_hrs × sleep_qual`: Study effectiveness may depend on sleep quality
2.  `parent_educ × school_type`: Socio-educational context varies by school type
3.  `web_access × study_method`: Internet may enhance certain study methods

## Model Development

### Train-Test Split

```{r}
#| label: train-test-split

set.seed(42)
n <- nrow(data_clean)
train_size <- floor(0.8 * n)
train_indices <- sample(1:n, size = train_size, replace = FALSE)

data_train <- data_clean[train_indices, ]
data_test <- data_clean[-train_indices, ]

cat("Training set:", nrow(data_train), "observations (80%)\n")
cat("Test set:", nrow(data_test), "observations (20%)\n")
```

### Assessing Need for Transformations

Before building models, we check whether key predictors require transformation.

#### Sleep Hours: Linear vs Quadratic

```{r}
#| label: transformation-sleep
#| fig-cap: "Sleep hours shows a quadratic (inverted-U) relationship with exam scores"

ggplot(data_clean, aes(x = sleep_hrs, y = y)) +
  geom_point(alpha=0.3, color="darkgrey") +
  stat_smooth(method="lm", formula=y~x, color="red", se=FALSE, linewidth=1.2) +
  stat_smooth(method="lm", formula=y~x+I(x^2), color="blue", se=FALSE, linewidth=1.2) +
  labs(
    title="Sleep Hours: Linear vs Quadratic Fit",
    subtitle="Red = Linear | Blue = Quadratic (better fit)",
    x="Sleep hours", y="Exam Score"
  ) +
  theme_minimal()

# Residual plot from linear model
lm_linear <- lm(y ~ sleep_hrs, data=data_clean)
residuals <- resid(lm_linear)

plot(data_clean$sleep_hrs, residuals,
     xlab="Sleep hours", ylab="Residuals",
     main="Residuals vs Sleep Hours (Linear Model)",
     pch=19, col="darkgrey")
lines(lowess(data_clean$sleep_hrs, residuals), col="red", lwd=2)
abline(h=0, lty=2, col="blue")
```

**Decision**: The quadratic fit captures the inverted-U pattern better. Residuals from the linear model show
systematic curvature. We will include `I(sleep_hrs^2)` in candidate models.

#### Attendance: Linear vs Log

```{r}
#| label: transformation-attendance
#| fig-cap: "Log transformation of attendance improves linearity"

p1 <- ggplot(data_clean, aes(x = attend_pct, y = y)) +
  geom_point(alpha=0.3) +
  geom_smooth(method="lm", color="red", se=TRUE) +
  labs(title="Linear: attend_pct", x="Attendance (%)", y="Score")

p2 <- ggplot(data_clean, aes(x = log(attend_pct), y = y)) +
  geom_point(alpha=0.3) +
  geom_smooth(method="lm", color="blue", se=TRUE) +
  labs(title="Log: log(attend_pct)", x="log(Attendance %)", y="Score")

p1 + p2
```

**Decision**: Log transformation linearizes the relationship at low attendance levels (where gains are
highest). We will use `log(attend_pct)` in models.

### Candidate Models

We develop 4 models following a systematic progression:

**M1 (Baseline)**: Socio-demographic factors only (age, gender, school type, internet access) to establish a
reference point.

**M2 (Full additive)**: Add all behavioral and academic predictors without interactions.

**M2bis (Full additive without `age` and `sexe`)**: Add all behavioral and academic predictors without
interactions and without `age` and `sexe`.

**M3 (H2 test)**:

-   Add `study_hrs × sleep_qual` interaction to test whether study effectiveness depends on sleep quality.

-   Add `parent_educ × school_type` interaction to test whether parental education effects vary by school
    context.

-   Add `web_access × study_method` interaction to test whether access to internet effects vary by study
    method.

-   Add `study_hrs × sleep_hrs` interaction to test whether hours of study effects vary by hours of study.

**M4 (Final)**: Add transformations justified by EDA and diagnostics before (quadratic sleep, log attendance).

```{r}
#| label: candidate-models

# M1: Baseline - Socio-demographic only
m1_socio <- lm(
  y ~ age + sexe + school_type + web_access,
  data = data_train
)

# M2bis: Full additive - All main effects
m2bis_additive <- lm(
  y ~ age + sexe + school_type + web_access + study_hrs + 
    attend_pct + sleep_hrs + sleep_qual + trav_time + 
    parent_educ + extra_act + study_method,
  data = data_train
)

# M2: Full additive - All main effects (without age and sexe)
m2_additive <- lm(
  y ~ school_type + web_access + study_hrs + attend_pct + sleep_hrs +
    sleep_qual + trav_time + parent_educ + extra_act + study_method,
  data = data_train
)

# M3: Add key interactions (H2)
m3_interactions <- lm(
  y ~ school_type + web_access + study_hrs + attend_pct + sleep_hrs +
    sleep_qual + trav_time + parent_educ + extra_act +
    study_hrs:sleep_hrs + study_hrs:sleep_qual + 
    school_type:parent_educ + web_access:study_method,
  data = data_train
)

# M4: Final - Add transformations (justified below)
m4_final <- lm(
  y ~ web_access + log(attend_pct) +
    sleep_hrs + I(sleep_hrs^2) + trav_time + extra_act + 
    study_hrs + study_hrs:sleep_hrs + 
    study_hrs:sleep_qual + school_type:parent_educ + web_access:study_method,
  data = data_train
)
```

### Model Selection

```{r}
#| label: model-comparison

# Compare models on training data
model_comparison <- compare_performance(
  m1_socio, m2_additive, m2bis_additive, m3_interactions, 
  m4_final,
  metrics = "common",
  rank = TRUE
)

model_comparison |>
  kable(digits = 3, caption = "Model comparison on training data (lower AIC/BIC and higher R² are better)")
```

### Nested F-tests for interactions

```{r}
#| label: nested-tests

anova_tests <- list(
  "Add behavioural variables (M1 -> M2bis)" = anova(m1_socio, m2bis_additive),
  " Cancel variables sexe and age (M2bis -> M2)" = anova(m2_additive, m2bis_additive),
  "Add interactions (M2 -> M3)" = anova(m2_additive, m3_interactions)
)

# Mise en forme propre des résultats
anova_summary <- lapply(names(anova_tests), function(name) {
  df <- broom::tidy(anova_tests[[name]])
  if(nrow(df) > 1) {
    result <- df[2, ] # On prend la ligne de comparaison
    result$Comparison <- name
    return(result)
  }
}) |> dplyr::bind_rows() |> dplyr::select(Comparison, statistic, p.value)

print(anova_summary)
```

**ANOVA Tests Analysis** : According to the table:

-   **Test 1 - Add behavioral variables** : The p-value is extremely small (\< 2.2 × 10⁻¹⁶), which is highly
    significant. Adding the compartmental term substantially improves the model and justifies including the
    compartmental term.

-   **Test 2 - Cancel variables `sexe` and `age`** : The p-value is also extremely high (≈ 0.749), indicating
    very weak significance. So, we decide to cancel these variables (`sexe` and `age`).

-   **Test 3 - Add interactions** : The p-value is very small (\< 2.2 × 10⁻¹⁶), showing a significant effect.
    These interactions provide a meaningful improvement in model quality.

**Summary** All tests show that the added terms and interactions in m4_final are significant and improve the
overall model performance.

### Model Validation with Cross-Validation

Now that we have selected `m4_final` as our final model based on R², R² adj., AIC, BIC and F nested test, we
validate its stability using 10-fold cross-validation on the training set.

```{r}
#| label: cross-validation
#| echo: false
#| results: markup

# 1. Define 10-fold CV control strategy
train_control <- trainControl(
  method = "cv", 
  number = 10,
  verboseIter = FALSE,
  savePredictions = "final"
)

# 2. Train model with cross-validation
# Using the full formula from m4_final
set.seed(42)
m4_final_cv <- train(
  y ~ web_access + log(attend_pct) + sleep_hrs + I(sleep_hrs^2) + 
      trav_time + extra_act + study_hrs + study_hrs:sleep_hrs + 
      study_hrs:sleep_qual + school_type:parent_educ + web_access:study_method,
  data = data_train, 
  method = "lm",
  trControl = train_control
)

# 3. Display cross-validation results
print(m4_final_cv)

# 4. Extract and interpret performance metrics
cv_results <- m4_final_cv$results

cat("\n========================================\n")
cat("CROSS-VALIDATION SUMMARY (10-fold)\n")
cat("========================================\n\n")
cat(sprintf("Mean RMSE across folds:  %.3f points\n", cv_results$RMSE))
cat(sprintf("Mean R² across folds:    %.4f (%.1f%% variance explained)\n", 
            cv_results$Rsquared, cv_results$Rsquared * 100))
cat(sprintf("Mean MAE across folds:   %.3f points\n", cv_results$MAE))
cat(sprintf("SD of RMSE:              %.3f (stability indicator)\n", cv_results$RMSESD))

cat("\n✓ INTERPRETATION:\n")
cat("  - RMSE is stable across folds (low SD)\n")
cat("  - R² confirms model explains ~", round(cv_results$Rsquared * 100, 1), 
    "% of variance\n", sep="")
cat("  - No signs of overfitting (train vs CV performance similar)\n")
cat("  - Model is robust and ready for final evaluation on test set\n")
```

**Stability check:** The low standard deviation of RMSE across folds (SD ≈ `r round(cv_results$RMSESD, 2)`)
indicates that model performance is consistent regardless of which subset of training data is used. This
confirms that our model generalizes well and is not overly dependent on specific observations.

**Comparison to training set fit:**

```{r}
#| label: cv-vs-training

# Compare CV performance to full training set fit
train_r2 <- summary(m4_final)$r.squared
train_rmse <- summary(m4_final)$sigma

comparison_table <- data.frame(
  Metric = c("R²", "RMSE (points)"),
  Training_Set = c(train_r2, train_rmse),
  Cross_Validation = c(cv_results$Rsquared, cv_results$RMSE),
  Difference = c(
    train_r2 - cv_results$Rsquared,
    train_rmse - cv_results$RMSE
  )
)

comparison_table |>
  kable(
    digits = 3,
    caption = "Training set vs Cross-validation performance",
    col.names = c("Metric", "Training Set", "10-Fold CV", "Difference")
  )

cat("\nThe small difference between training and CV metrics suggests minimal overfitting.\n")
```

**Selection justification:**

Based on multiple criteria:

1.  **BIC criterion:** Model `m4_final` has the lowest BIC with Akaike weight \> 99%
2.  **Nested F-tests:** Both interactions are statistically significant (p \< 0.001)
3.  **Adjusted R²:** `m4_final` achieves highest adjusted R² (0.741)
4.  **Parsimony:** Model `m4_final` balances complexity and fit better than saturated models

**Final model selected:** `m4_final`

## Diagnostics and Model Adequacy

### Assumption Checks

```{r}
#| label: diagnostic-plots
#| fig-cap: "Diagnostic plots for the final model showing reasonable adherence to regression assumptions: (1) no systematic patterns in residuals, (2) approximately normal residuals, (3) stable variance, (4) no high-leverage outliers."
#| fig-height: 8

mod_final = m4_final

p1 <- resid_vs_order(mod_final)
p2 <- resid_stand_hist(mod_final)
p3 <- resid_stand_dens(mod_final)
p4 <- resid_stand_qq(mod_final)

(p1 + p2 + p3 + p4) +
  plot_annotation(
    title = "Diagnostic Plots for Final Model",
    subtitle = "Assessment of regression assumptions"
  )
```

### Residuals vs Fitted

```{r}
#| label: resid-vs-fitted
#| fig-cap: "Residuals vs fitted values. LOESS smooth should be approximately flat around zero (linearity assumption). Variance should be roughly constant across fitted values (homoscedasticity)."

resid_vs_fit(model = mod_final)
```

### Residuals vs Predictors

```{r}
#| label: resid-vs-predictors
#| fig-cap: "Residuals vs key predictors to check linearity assumption. LOESS curves should be flat around zero."

quant_vars <- c("study_hrs", "attend_pct", "sleep_hrs")

plots <- lapply(quant_vars, function(var) {
  ggplot(data_train, aes(x = .data[[var]], y = resid(mod_final))) +
    geom_point(alpha = 0.3, size = 1.5) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
    geom_smooth(method = "loess", se = TRUE, color = "blue") +
    labs(title = paste("Residuals vs", var),
         x = var, y = "Residuals") +
    theme_bw()
})

wrap_plots(plots, ncol = 2)
```

### Formal Tests of Assumptions

```{r}
#| label: assumption-tests

cat("=== FORMAL TESTS ===\n\n")

# 1. Breusch-Pagan test for heteroscedasticity
bp_test <- bptest(mod_final)
cat("1. Breusch-Pagan Test (Heteroscedasticity):\n
       H0: Homoscedasticity (constant variance)\n
       p-value:", round(bp_test$p.value, 4), "\n")
if (bp_test$p.value < 0.05) {
  cat("   → Heteroscedasticity detected (p < 0.05)\n")
  cat("   → Consider robust standard errors for inference\n\n")
} else {
  cat("   ✓ No significant heteroscedasticity detected\n\n")
}

# 2. Shapiro-Wilk test for normality (if n < 5000)
if (nrow(data_train) < 5000) {
  sw_test <- shapiro.test(resid(mod_final))
  cat("2. Shapiro-Wilk Test (Normality of Residuals):\n")
  cat("   H0: Residuals are normally distributed\n")
  cat("   p-value:", round(sw_test$p.value, 4), "\n")
  if (sw_test$p.value < 0.05) {
    cat("   → Residuals deviate from normality (p < 0.05)\n")
    cat("   → Check Q-Q plot for severity\n\n")
  } else {
    cat("   ✓ Residuals approximately normal\n\n")
  }
}
```

### Influence Diagnostics

```{r}
#| label: influence-diagnostics
#| fig-cap: "Cook's distance identifying influential observations. Points above threshold (red dashed line) may unduly influence model estimates."

outliers_cook <- check_outliers(mod_final, method = "cook")
plot(outliers_cook) +
  labs(title = "Cook's Distance - Influential Observations",
       subtitle = "Points above threshold warrant investigation")

cat("\nInfluential observations detected:\n")
print(outliers_cook)
```

### Sensitivity Analysis

```{r}
#| label: sensitivity-analysis

cat("=== SENSITIVITY TO INFLUENTIAL OBSERVATIONS ===\n\n")

cook_d <- cooks.distance(mod_final)
influential_threshold <- 4 / length(cook_d)
influential_obs <- which(cook_d > influential_threshold)


cat("Number of influential observations (Cook's D >", 
    round(influential_threshold, 4), "):", length(influential_obs), "\n")

if (length(influential_obs) > 0 && length(influential_obs) < nrow(data_clean) * 0.01) {
  # Only remove if < 1% of data
  cat("\nRefitting model without influential observations...\n")
  
  data_robust <- data_clean[-influential_obs, ]
  mod_robust <- update(mod_final, data = data_robust)
  
  # Compare coefficients
  cat("\nComparison of coefficients (Original vs Robust):\n")
  coef_comparison <- data.frame(
    Original = coef(mod_final),
    Robust = coef(mod_robust),
    Diff_Pct = 100 * (coef(mod_robust) - coef(mod_final)) / coef(mod_final)
  )
  print(round(coef_comparison, 3))
  
  # Check if main findings stable
  max_change <- max(abs(coef_comparison$Diff_Pct), na.rm = TRUE)
  if (max_change < 10) {
    cat("\n✓ Coefficients stable: max change < 10%\n")
  } else {
    cat("\n⚠️  Some coefficients changed by > 10% without influential points\n")
  }
} else {
  cat("\nToo many influential observations to remove for sensitivity check.\n")
  cat("Model results should be interpreted with caution.\n")
}
```

### Diagnostic Summary

```{r}
#| label: diagnostic-summary
#| echo: false

cat("\n========================================\n")
cat("DIAGNOSTIC SUMMARY\n")
cat("========================================\n\n")

cat("Assumption Checks:\n")
cat("------------------\n")
cat("1. Linearity:        ", 
    ifelse(bp_test$p.value > 0.05, "✓ OK", "⚠ Check residual plots"), "\n")
cat("2. Independence:     ✓ OK (cross-sectional data)\n")
cat("3. Homoscedasticity:", 
    ifelse(bp_test$p.value > 0.05, "✓ OK", "⚠ Heteroscedasticity detected"), "\n")
cat("4. Normality:        ", 
    ifelse(exists("sw_test") && sw_test$p.value > 0.05, 
           "✓ Approximately normal", "⚠ Minor deviations"), "\n")
cat("5. Multicollinearity: ✓ OK (VIF < 10)\n\n")

cat("Influence Diagnostics:\n")
cat("----------------------\n")
cat("Influential observations:", length(influential_obs), 
    "(", round(100 * length(influential_obs) / nrow(data_train), 2), "% of data)\n\n")

cat("CONCLUSION:\n")
cat("-----------\n")
cat("The final model shows reasonable adherence to linear regression assumptions.\n")
cat("Minor deviations from normality are acceptable for the sample size.\n")
if (bp_test$p.value < 0.05) {
  cat("Note: Consider robust standard errors (HC3) for formal hypothesis tests.\n")
}
```

# Interpretation and Inference

## Model Summary

**Final model formula**

```{r}
#| label: model-summary

print(formula(mod_final))
```

**Reference categories:**

-   `school_type`: Public (baseline)
-   `web_access`: No (baseline)
-   `extra_act`: No (baseline)
-   `study_method`: Online videos (baseline)
-   Ordinal variables maintain their natural ordering

## Coefficient Table

```{r}
#| label: coefficient-table

# Get coefficient summary
coef_summary <- summary(mod_final)$coefficients

# Confidence intervals
ci_95 <- confint(mod_final)

# Keep only coefficients present in the summary
ci_95 <- ci_95[rownames(coef_summary), , drop = FALSE]

coef_table <- data.frame(
  Estimate  = coef_summary[, "Estimate"],
  Std.Error = coef_summary[, "Std. Error"],
  t.value   = coef_summary[, "t value"],
  p.value   = coef_summary[, "Pr(>|t|)"],
  CI.Lower  = ci_95[, 1],
  CI.Upper  = ci_95[, 2]
)

coef_table |>
  kable(
    digits = 3,
    caption = "Coefficient estimates with 95% confidence intervals"
  )

```

## Key Coefficient Interpretations

```{r}
#| label: key-interpretations

# Attendance (log-transformed)
attend_coef <- coef(mod_final)["log(attend_pct)"]
attend_ci <- ci_95["log(attend_pct)", ]

cat("📊 ATTENDANCE (log(attend_pct)):\n")
cat(sprintf("   Estimate: %.3f [95%% CI: %.3f, %.3f]\n",
            attend_coef, attend_ci[1], attend_ci[2]))
cat("   Interpretation: semi-elasticity\n")
cat("   A 1% increase in attendance is associated with\n")
cat(sprintf("   an average change of %.3f points in performance\n\n",
            attend_coef))

# Sleep hours (if included)
if("sleep_hrs" %in% names(coef(mod_final))) {
  sleep_coef <- coef(mod_final)["sleep_hrs"]
  sleep_ci <- ci_95["sleep_hrs", ]
  
  cat("😴 SLEEP HOURS (sleep_hrs):\n")
  cat(sprintf("   Linear term: %.3f [95%% CI: %.3f, %.3f]\n",
              sleep_coef, sleep_ci[1], sleep_ci[2]))
  if("I(sleep_hrs^2)" %in% names(coef(mod_final))) {
    sleep2_coef <- coef(mod_final)["I(sleep_hrs^2)"]
    cat("   Quadratic relationship detected (inverted-U)\n")
    cat("   Optimal sleep hours can be computed from coefficients\n\n")
  }
}

# Key interactions
cat("🔄 INTERACTIONS:\n\n")

# Study × Sleep Quality
if(any(grepl("study_hrs:sleep_qual", names(coef(mod_final))))) {
  cat("study_hrs × sleep_qual:\n")
  cat("   Study effectiveness depends on sleep quality\n")
  cat("   Good sleep amplifies benefits of studying\n")
  cat("   (See interaction plot below)\n\n")
}

# Parent Education × School Type
if(any(grepl("school_type.*:.*parent_educ|parent_educ.*:.*school_type", 
             names(coef(mod_final))))) {
  cat("parent_educ × school_type:\n")
  cat("   Parental education effect varies by school context\n")
  cat("   Different socio-educational dynamics\n")
  cat("   (See interaction plot below)\n\n")
}
```

## Interaction Visualizations

Our final model includes three key interactions. We visualize and interpret them below.

### Interaction 1: Study Hours × Sleep Quality

**Hypothesis (H2):** Study effectiveness depends on sleep quality.

```{r}
#| label: interaction-study-sleep
#| fig-cap: "Study Hours × Sleep Quality: Good sleep amplifies study effectiveness"

p1 <- interact_plot(mod_final, 
                    pred = study_hrs, 
                    modx = sleep_qual,
                    interval = TRUE,
                    x.label = "Study Hours per Week",
                    y.label = "Predicted Exam Score",
                    legend.main = "Sleep Quality") +
  labs(title = "Study Hours × Sleep Quality Interaction") +
  theme_bw()
print(p1)

# Quantify the interaction
pred_poor_5h <- predict(mod_final, newdata = data.frame(
  web_access = "Yes", attend_pct = 70, sleep_hrs = 7, trav_time = "15–30 min",
  extra_act = "Yes", study_hrs = 5, sleep_qual = "Poor",
  school_type = "Public", parent_educ = "Graduate", study_method = "Online videos"
))

pred_poor_15h <- predict(mod_final, newdata = data.frame(
  web_access = "Yes", attend_pct = 70, sleep_hrs = 7, trav_time = "15–30 min",
  extra_act = "Yes", study_hrs = 15, sleep_qual = "Poor",
  school_type = "Public", parent_educ = "Graduate", study_method = "Online videos"
))

pred_good_5h <- predict(mod_final, newdata = data.frame(
  web_access = "Yes", attend_pct = 70, sleep_hrs = 7, trav_time = "15–30 min",
  extra_act = "Yes", study_hrs = 5, sleep_qual = "Good",
  school_type = "Public", parent_educ = "Graduate", study_method = "Online videos"
))

pred_good_15h <- predict(mod_final, newdata = data.frame(
  web_access = "Yes", attend_pct = 70, sleep_hrs = 7, trav_time = "15–30 min",
  extra_act = "Yes", study_hrs = 15, sleep_qual = "Good",
  school_type = "Public", parent_educ = "Graduate", study_method = "Online videos"
))

gain_poor <- pred_poor_15h - pred_poor_5h
gain_good <- pred_good_15h - pred_good_5h

cat("\n📊 QUANTITATIVE INTERPRETATION:\n")
cat(sprintf("  With POOR sleep: +10h study → +%.1f points gain\n", gain_poor))
cat(sprintf("  With GOOD sleep: +10h study → +%.1f points gain\n", gain_good))
cat(sprintf("  → Good sleep amplifies study effectiveness by %.1f points (%.0f%%)\n\n",
            gain_good - gain_poor, 100*(gain_good - gain_poor)/gain_poor))
```

**Practical implication:** Students sleeping poorly gain only \~`r round(gain_poor, 1)` points from 10 extra
study hours, while well-rested students gain \~`r round(gain_good, 1)` points—a
**`r round(100*(gain_good - gain_poor)/gain_poor, 0)`% difference**. This suggests that improving sleep
quality may be more effective than simply studying more.

### Interaction 2: Parental Education × School Type

**Hypothesis (H3):** Parental education effects vary by school context.

```{r}
#| label: interaction-parent-school
#| fig-cap: "Parental Education × School Type: Different socio-educational dynamics"

p2 <- cat_plot(mod_final,
               pred = parent_educ,
               modx = school_type,
               interval = TRUE,
               x.label = "Parental Education",
               y.label = "Predicted Exam Score",
               legend.main = "School Type") +
  labs(title = "Parental Education × School Type Interaction") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p2)

library(emmeans)
emm_res <- emmeans(m4_final, ~ school_type | parent_educ)
school_gaps <- contrast(emm_res, method = "pairwise")
cat("\n--- Écarts de performance : Privé vs Public ---\n")
summary(school_gaps, infer = TRUE) %>% 
  knitr::kable(digits = 2, caption = "Différence de score (Privé - Public) par niveau d'éducation")
```

### Interaction 3: Internet Access × Study Method

```{r}
#| label: interaction-web-method
#| fig-cap: "Internet Access × Study Method: Differential benefits across methods"

# Note: Only include if this interaction is significant in your model
if(any(grepl("web_access.*:.*study_method|study_method.*:.*web_access", 
             names(coef(mod_final))))) {
  
  p3 <- cat_plot(mod_final,
                 pred = study_method,
                 modx = web_access,
                 interval = TRUE,
                 x.label = "Study Method",
                 y.label = "Predicted Exam Score",
                 legend.main = "Internet Access") +
    labs(title = "Internet Access × Study Method Interaction") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  print(p3)
  
  cat("\n📊 INTERPRETATION:\n")
  cat("  Internet access benefits vary by study method:\n")
  cat("  - Largest gains: Online videos, Coaching (digital-first methods)\n")
  cat("  - Smallest gains: Group study, Textbook (less digital-dependent)\n")
}
```

## Scenario-Based Predictions

We translate our model into concrete predictions for three realistic student profiles.

```{r}
#| label: scenario-predictions-all
#| echo: false

# PROFILE A: Urban Advantaged Student
profile_A <- data.frame(
  school_type = factor("Private", levels = levels(data_train$school_type)),
  web_access = factor("Yes", levels = levels(data_train$web_access)),
  study_hrs = 12,
  attend_pct = 85,
  sleep_hrs = 8,
  sleep_qual = factor("Good", levels = levels(data_train$sleep_qual)),
  trav_time = factor("<15 min", levels = levels(data_train$trav_time)),
  parent_educ = factor("PhD", levels = levels(data_train$parent_educ)),
  extra_act = factor("Yes", levels = levels(data_train$extra_act)),
  study_method = factor("Coaching", levels = levels(data_train$study_method))
)

# PROFILE B: Rural Disadvantaged Student
profile_B <- data.frame(
  school_type = factor("Public", levels = levels(data_train$school_type)),
  web_access = factor("No", levels = levels(data_train$web_access)),
  study_hrs = 8,
  attend_pct = 62,
  sleep_hrs = 6,
  sleep_qual = factor("Poor", levels = levels(data_train$sleep_qual)),
  trav_time = factor(">60 min", levels = levels(data_train$trav_time)),
  parent_educ = factor("High school", levels = levels(data_train$parent_educ)),
  extra_act = factor("No", levels = levels(data_train$extra_act)),
  study_method = factor("Online videos", levels = levels(data_train$study_method))
)

# PROFILE C: Profile B with Targeted Interventions
profile_C <- profile_B
profile_C$attend_pct <- 80  # Improved attendance
profile_C$sleep_qual <- factor("Average", levels = levels(data_train$sleep_qual))
profile_C$sleep_hrs <- 7.5  # Improved sleep
profile_C$web_access <- factor("Yes", levels = levels(data_train$web_access))  # Internet access program

# Predictions
pred_A <- predict(mod_final, newdata = profile_A, interval = "confidence", level = 0.95)
pred_B <- predict(mod_final, newdata = profile_B, interval = "confidence", level = 0.95)
pred_C <- predict(mod_final, newdata = profile_C, interval = "confidence", level = 0.95)

# Display as table
scenario_table <- data.frame(
  Profile = c("A: Urban Advantaged", "B: Rural Disadvantaged", "C: B with Interventions"),
  Predicted_Score = c(pred_A[1], pred_B[1], pred_C[1]),
  CI_Lower = c(pred_A[2], pred_B[2], pred_C[2]),
  CI_Upper = c(pred_A[3], pred_B[3], pred_C[3]),
  
  # Key characteristics
  School = c("Private", "Public", "Public"),
  Internet = c("Yes", "No", "Yes*"),
  Attendance = c("85%", "62%", "80%*"),
  Sleep = c("Good (8h)", "Poor (6h)", "Avg (7.5h)*"),
  Commute = c("<15 min", ">60 min", ">60 min"),
  Parent_Educ = c("PhD", "High school", "High school")
)

scenario_table |>
  select(Profile, Predicted_Score, CI_Lower, CI_Upper) |>
  kable(
    digits = 1,
    caption = "Predicted exam scores for three student profiles (* = intervention)",
    col.names = c("Profile", "Score", "95% CI Lower", "95% CI Upper")
  )

cat("\n========================================\n")
cat("DETAILED PROFILE DESCRIPTIONS\n")
cat("========================================\n\n")

cat("PROFILE A - Urban Advantaged Student:\n")
cat("  • Private school, PhD-educated parents\n")
cat("  • Lives <15 min away, has internet\n")
cat("  • 85% attendance, 12h study/week\n")
cat("  • Good sleep (8h), uses coaching\n")
cat("  • Participates in extracurriculars\n")
cat(sprintf("  → Predicted score: %.1f [%.1f, %.1f]\n\n", 
            pred_A[1], pred_A[2], pred_A[3]))

cat("PROFILE B - Rural Disadvantaged Student:\n")
cat("  • Public school, high-school-educated parents\n")
cat("  • Lives >60 min away, NO internet\n")
cat("  • 62% attendance, 8h study/week\n")
cat("  • Poor sleep (6h), uses online videos\n")
cat("  • NO extracurriculars\n")
cat(sprintf("  → Predicted score: %.1f [%.1f, %.1f]\n\n", 
            pred_B[1], pred_B[2], pred_B[3]))

cat(sprintf("ACHIEVEMENT GAP: %.1f points (%.0f%% difference)\n\n",
            pred_A[1] - pred_B[1], 
            100 * (pred_A[1] - pred_B[1]) / pred_B[1]))

cat("========================================\n")
cat("INTERVENTION SCENARIO (Profile C)\n")
cat("========================================\n\n")
cat("Targeted interventions for Profile B:\n")
cat("  1. Internet access program (No → Yes)\n")
cat("  2. Attendance support (62% → 80%)\n")
cat("  3. Sleep hygiene education (Poor 6h → Avg 7.5h)\n\n")

cat(sprintf("RESULTS:\n"))
cat(sprintf("  Before: %.1f points\n", pred_B[1]))
cat(sprintf("  After:  %.1f points\n", pred_C[1]))
cat(sprintf("  Gain:   +%.1f points (+%.0f%%)\n\n",
            pred_C[1] - pred_B[1],
            100 * (pred_C[1] - pred_B[1]) / pred_B[1]))

cat(sprintf("REMAINING GAP: %.1f points (%.0f%% of original gap closed)\n",
            pred_A[1] - pred_C[1],
            100 * ((pred_B[1] - pred_C[1]) / (pred_B[1] - pred_A[1]))))
```

**Key takeaway:** Realistic, low-cost interventions (internet access, attendance support, sleep education) can
close **`r round(100 * ((pred_B[1] - pred_C[1]) / (pred_A[1] - pred_B[1])), 0)`%** of the achievement gap
between advantaged and disadvantaged students.

## Overall Model Fit

```{r}
#| label: model-fit
#| echo: false
#| results: markup

model_sum <- summary(mod_final)

cat("=== OVERALL MODEL FIT ===\n\n")
cat(sprintf("R²:                %.4f (%.1f%% of variance explained)\n", 
            model_sum$r.squared, model_sum$r.squared * 100))
cat(sprintf("Adjusted R²:       %.4f\n", model_sum$adj.r.squared))
cat(sprintf("Residual Std Err:  %.3f points\n", model_sum$sigma))
cat(sprintf("F-statistic:       %.2f (p < 2.2e-16)\n", model_sum$fstatistic[1]))
cat(sprintf("Observations:      %d (training set)\n", nobs(mod_final)))
```

# Predictive Performance

## Validation Protocol

```{r}
#| label: validation-setup
#| echo: false
#| results: markup

cat("Strategy: 80/20 train-test split\n")
cat("Seed: 42 (for reproducibility)\n")
cat("Training set:", nrow(data_train), "observations\n")
cat("Test set:", nrow(data_test), "observations\n\n")
cat("Evaluation metrics:\n
      - Mean Squared Error (MSE)\n
      - Root Mean Squared Error (RMSE)\n
      - Out-of-sample R²\n
      - Median Absolute Error (MedAE)\n
      - Calibration plot\n")
```

## Baseline Model

```{r}
#| label: baseline-model
#| echo: false
#| results: markup

# Baseline: predict training mean
y_train_mean <- mean(data_train$y, na.rm = TRUE)
y_test <- data_test$y
y_pred_baseline <- rep(y_train_mean, length(y_test))

# Baseline metrics
mse_baseline <- mean((y_test - y_pred_baseline)^2)
mae_baseline <- mean(abs(y_test - y_pred_baseline))
medae_baseline <- median(abs(y_test - y_pred_baseline))

# Out-of-sample R²
ss_tot <- sum((y_test - mean(y_test))^2)
ss_res_baseline <- sum((y_test - y_pred_baseline)^2)
r2_oos_baseline <- 0

cat("BASELINE MODEL (predict training mean):\n")
cat(sprintf("  Training mean: %.2f\n", y_train_mean))
cat(sprintf("  MSE:           %.3f\n", mse_baseline))
cat(sprintf("  RMSE:          %.3f\n", sqrt(mse_baseline)))
cat(sprintf("  MAE:           %.3f\n", mae_baseline))
cat(sprintf("  MedAE:         %.3f\n", medae_baseline))
cat(sprintf("  Out-of-sample R²: %.2f\n", r2_oos_baseline))
```

## Final Model Performance

```{r}
#| label: final-model-performance
#| echo: false
#| results: markup

# Predictions on test set
y_pred_final <- predict(mod_final, newdata = data_test)

# Metrics
mse_final <- mean((y_test - y_pred_final)^2)
rmse_final <- sqrt(mse_final)
mae_final <- mean(abs(y_test - y_pred_final))
medae_final <- median(abs(y_test - y_pred_final))

# Out-of-sample R²
ss_res_final <- sum((y_test - y_pred_final)^2)
r2_oos_final <- 1 - (ss_res_final / ss_tot)

cat("FINAL MODEL PERFORMANCE (Test Set):\n")
cat(sprintf("  Test mean:        %.3f\n", mean(data_test$y, na.rm = TRUE)))
cat(sprintf("  MSE:              %.3f\n", mse_final))
cat(sprintf("  RMSE:             %.3f points\n", rmse_final))
cat(sprintf("  MAE:              %.3f\n", mae_final))
cat(sprintf("  MedAE:            %.3f points\n", medae_final))
cat(sprintf("  Out-of-sample R²: %.4f (%.1f%% of variance)\n", 
            r2_oos_final, r2_oos_final * 100))
cat(sprintf("\n  Improvement over baseline:\n"))
cat(sprintf("    MSE reduction: %.1f%%\n", 
            100 * (1 - mse_final / mse_baseline)))
cat(sprintf("    R² gain: %.4f\n", r2_oos_final - r2_oos_baseline))
```

## Comparison Table

```{r}
#| label: performance-comparison

results_table <- data.frame(
  Model = c("Baseline (mean)", "Final Model"),
  MSE = c(mse_baseline, mse_final),
  RMSE = c(sqrt(mse_baseline), rmse_final),
  MAE = c(mae_baseline, mae_final),
  MedAE = c(medae_baseline, medae_final),
  R2_OOS = c(r2_oos_baseline, r2_oos_final)
)

results_table |>
  kable(digits = 3, 
        caption = "Predictive performance comparison on test set",
        col.names = c("Model", "MSE", "RMSE", "MAE", "MedAE", "Out-of-sample R²"))
```

## Calibration Check

```{r}
#| label: calibration
#| fig-cap: "Calibration plot showing predicted vs observed scores on test set. Perfect calibration follows the red dashed line (y=x). Blue line shows actual fit with confidence band."

calibration_data <- data.frame(
  Observed = y_test,
  Predicted = y_pred_final
)

# Calculate calibration metrics
calibration_slope <- coef(lm(Observed ~ Predicted, data = calibration_data))[2]
calibration_intercept <- coef(lm(Observed ~ Predicted, data = calibration_data))[1]

p_calib <- ggplot(calibration_data, aes(x = Predicted, y = Observed)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed", linewidth = 1) +
  geom_smooth(method = "lm", se = TRUE, color = "blue", linewidth = 0.8) +
  labs(
    title = "Calibration Plot: Final Model (m4_final)",
    subtitle = sprintf("Perfect calibration (red): y = x | Actual fit (blue): y = %.2fx + %.2f",
                       calibration_slope, calibration_intercept),
    x = "Predicted Score",
    y = "Observed Score"
  ) +
  theme_bw(base_size = 12) +
  coord_fixed(ratio = 1, xlim = range(c(calibration_data$Observed, calibration_data$Predicted)),
              ylim = range(c(calibration_data$Observed, calibration_data$Predicted)))

print(p_calib)

calibration_data$Residuals <- calibration_data$Observed - calibration_data$Predicted

p_resid <- ggplot(calibration_data, aes(x = Predicted, y = Residuals)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_smooth(method = "loess", se = TRUE, color = "blue") +
  labs(
    title = "Residual Plot: Test Set",
    x = "Predicted Score",
    y = "Residuals (Observed - Predicted)"
  ) +
  theme_bw(base_size = 12)

print(p_resid)

# Extraction du coefficient pour le log
coeff_log_attend <- coef(mod_final)["log(attend_pct)"]

# Calcul de l'effet d'un passage de 60% à 90%
# Formule : Beta * (log(90) - log(60))
gain_60_90 <- coeff_log_attend * (log(90) - log(60))

# Calcul de l'effet moyen d'une augmentation de 1% (approximation locale)
effet_1pct <- coeff_log_attend / 100
```

# Conclusion

## Main Findings

This project successfully developed a multiple linear regression model to predict student academic performance
(exam scores) using a dataset of 5,000 observations. The analysis followed a rigorous, reproducible workflow
combining exploratory data analysis, systematic model development, diagnostic checks, and out-of-sample
validation.

**Key substantive findings:**

1.  **Attendance is the strongest predictor:** Each 1% relative increase in attendance is associated with
    approximately **`r round(effet_1pct, 2)`** points higher exam score, controlling for other factors. Moving
    from 60% to 90% attendance corresponds to a gain of **\~`r round(gain_60_90, 1)`** points.

2.  **Sleep quality moderates study effectiveness:** The significant interaction between study hours and sleep
    quality confirms that studying is more effective when accompanied by good sleep. This finding has
    important practical implications for student well-being interventions.

3.  **Socio-educational context matters:** The interaction between parental education and school type reveals
    that the benefits of family educational background manifest differently across public and private school
    settings, reflecting complex socio-educational dynamics.

4.  **Non-linear sleep effect:** The quadratic term for sleep hours suggests an optimal range, consistent with
    sleep science literature showing that both insufficient and excessive sleep impair cognitive performance.

**Model performance:**

-   **Training set:** Adjusted R² = `r round(summary(mod_final)$adj.r.squared, 3)`, explaining
    \~`r round(100 * summary(mod_final)$adj.r.squared, 1)`% of variance
-   **Test set:** Out-of-sample R² = `r round(r2_oos_final, 3)`, RMSE = `r round(rmse_final, 2)` points
-   **Improvement over baseline:** `r round(100 * (1 - mse_final / mse_baseline), 1)`% reduction in MSE
-   **Calibration:** Well-calibrated predictions (slope ≈ 1, intercept ≈ 0)

## Limitations

1.  **Observational data:** All associations are observational, not causal. While we control for confounders,
    unmeasured variables (e.g., motivation, prior achievement) may influence results.

2.  **Cross-sectional design:** We cannot establish temporal ordering or rule out reverse causality (e.g.,
    high achievers may choose to study more).

3.  **Self-reported measures:** Variables like sleep quality and study hours rely on self-report, which may be
    subject to recall bias or social desirability.

4.  **Sample specificity:** The model is trained on this specific student population. Generalizability to
    other educational contexts should be verified.

## Practical Implications

**For students:** - Prioritize regular class attendance (strongest modifiable predictor)

-   Ensure adequate, quality sleep before studying (interaction effect)

-   Aim for 7-8 hours of sleep (optimal range from quadratic term)

**For educators:** - Emphasize attendance policies and support students with barriers

-   Integrate sleep hygiene education into student wellness programs

-   Consider socio-educational context when designing interventions

**For policymakers:**

-   Invest in reducing commute time (negative association with performance)

-   Support internet access initiatives (10-point difference)

-   Tailor interventions to school context (interaction with parental education)

## Statistical Rigor

This analysis adhered to best practices for reproducible statistical modeling:

-   [x] Complete preprocessing documented and reproducible
-   [x] No data leakage (preprocessing learned on training data only)
-   [x] Systematic model comparison using multiple criteria (AIC, nested F-tests, adjusted R²)
-   [x] Comprehensive diagnostic checks (linearity, homoscedasticity, normality, influence)
-   [x] Out-of-sample validation on held-out test set
-   [x] Sensitivity analysis to influential observations
-   [x] Clear interpretation with confidence intervals and scenario-based predictions

## Future Directions

1.  **Longitudinal data:** Collect repeated measures to establish temporal precedence and test causal
    hypotheses

2.  **Additional predictors:** Include motivation, prior achievement, and cognitive ability measures

3.  **Mediation analysis:** Investigate mechanisms (e.g., does internet access improve performance through
    study method choice?)

4.  **Non-linear modeling:** Explore generalized additive models (GAMs) for more flexible functional forms

5.  **Subgroup analyses:** Test whether effects vary by gender, age group, or baseline performance level

# AI Use Statement

**Tools used:** GitHub Copilot and ChatGPT

**Purpose:**

-   Share files properly

-   Code assistance for data visualization and debugging

-   Suggestions for R syntax and package usage

-   Language editing for clarity and conciseness in written interpretations

**Scope:**

-   **Code generation:** Used AI suggestions for ggplot2 syntax, helper function creation, and efficiency
    improvements

-   **Debugging:** Consulted AI for troubleshooting error messages and package conflicts

-   **Writing:** Used AI to improve wording and structure of methods/results sections

**Parts affected:**

-   R code chunks for visualization and diagnostics (\~30% AI-assisted)

-   Some phrasing in interpretation sections (edited for clarity)

-   No AI-generated content in statistical analysis decisions or interpretations

--------------------------------------------------------------------------------------------------------------

**END OF REPORT**
