---
title: "Linear Models Project in R"
subtitle: "Predicting Student Academic Performance"
author:
  - name: "BEN SAID Daniel ; DESCHILDRE Mateo ; ROY Damien"
    affiliation: "M1–MIDO, Université Paris Dauphine-PSL"
date: last-modified
lang: en
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: body
    toc-title: "Contents"
    number-sections: true
    smooth-scroll: true
    page-layout: full
    linestretch: 1.2
    fontsize: 125%
    self-contained: true
    df-print: default
    code-fold: true
    code-tools: true
    code-line-numbers: false
    code-overflow: scroll
    code-copy: true
    highlight-style: github
    theme: cosmo
    code-block-bg: "#F5F5F5"
    html-math-method: mathjax
    fig-format: retina
    fig-width: 13
    fig-height: 7
    fig-align: center
    fig-pos: "H"
    fig-cap-location: bottom
    mainfont: "Source Sans Pro"
execute:
  echo: true
  warning: false
  message: false
  error: false
editor:
  markdown:
    wrap: 110
    canonical: true
knitr:
  opts_chunk:
    tidy: styler
    comment: ''
    dev: svg
---

```{r}
#| label: setup
#| include: false

# Reproducibility
set.seed(42)

# Global options
options(pillar.width = 130, width = 200, scipen = 999, digits = 5)

# Load packages
library("broom")
library("car")
library("caret")
library("collapse")
library("correlation")
library("datawizard")
library("effectsize")
library("GGally")
library("ggfortify")
library("ggpubr")
library("ggrepel")
library("glue")
library("gtsummary")
library("here")
library("insight")
library("janitor")
library("kableExtra")
library("lmtest")
library("marginaleffects")
library("matrixTests")
library("modelbased")
library("multcomp")
library("openxlsx")
library("parameters")
library("patchwork")
library("performance")
library("pROC")
library("qqplotr")
library("rstatix")
library("scales")
library("see")
library("tidyverse")
library("interactions")

source(here("script", "helper_functions.R"))
```

# Introduction

## Context and Research Problem

Academic performance is influenced by a complex interplay of factors including study habits, living
conditions, socio-educational context, and health-related behaviors. Understanding these relationships is
crucial for educators, policymakers, and students themselves to identify actionable interventions that can
improve educational outcomes.

This project aims to predict student academic performance (measured by exam scores) using multiple linear
regression, while exploring the relative importance of various predictors and potential interactions between
them. The analysis follows a dual objective: building an **interpretable model** for inference and
understanding relationships, and achieving **strong predictive performance** on unseen data.

## Dataset Overview

The dataset comprises **5,000 observations** collected from students, with **15 variables** covering different
dimensions of their academic journey and daily life:

**Life habits and well-being:**

-   `sleep_hrs` (continuous): nightly sleep duration
-   `sleep_qual` (ordinal): subjective sleep quality (Poor, Average, Good)
-   `trav_time` (ordinal): daily commute time to school

**Study habits:**

-   `study_hrs` (continuous): weekly study hours
-   `study_method` (categorical): preferred study method (6 categories)
-   `attend_pct` (continuous): percentage of class attendance

**Contextual and environmental factors:**

-   `age` (continuous): student age in years
-   `sexe` (categorical): gender (Female, Male, Other)
-   `school_type` (categorical): type of institution (Public, Private)
-   `web_access` (categorical): home internet access (Yes, No)
-   `parent_educ` (ordinal): parental education level (6 levels)
-   `extra_act` (categorical): participation in extracurricular activities (Yes, No)

The dependent variable `y` represents a quantitative measure of academic performance (exam score, range
0-100).

## Research Hypotheses

Based on existing literature in education science and cognitive psychology, we formulate the following
hypotheses:

**H1: Non-linear effect of sleep**\
Sleep duration (`sleep_hrs`) exhibits a quadratic (inverted-U) relationship with performance, where both
insufficient and excessive sleep negatively affect scores.

**H2: Interaction between study quantity and sleep quality**\
The effectiveness of study time (`study_hrs`) depends on sleep quality (`sleep_qual`). Good sleep should
amplify the benefits of studying.

**H3: Interaction between family background and school type**\
The effect of parental education (`parent_educ`) on performance varies by school type (`school_type`),
reflecting different socio-educational contexts.

**H4: Main effects**\
- Higher attendance (`attend_pct`) should positively predict performance - Internet access (`web_access`)
facilitates access to educational resources - Extracurricular activities (`extra_act`) may develop
transferable skills - Excessive commute time (`trav_time`) negatively impacts available study and rest time

# Methods

## Data Management and Preparation

### Data Import and Structure

```{r}
#| label: data-import

data <- read.csv(here("data", "project.csv"))

cat("Sample size:", nrow(data), "observations\n")
cat("Number of variables:", ncol(data), "\n\n")

glimpse(data)
```

### Variable Conversion and Labeling

We convert coded categorical variables into properly labeled factors with meaningful reference categories:

```{r}
#| label: data-cleaning

data_clean <- data |>
  mutate(
    sexe = factor(sexe, levels = c(1, 2, 3), labels = c("Female", "Male", "Other")),
    school_type = factor(school_type, levels = c(1, 2), labels = c("Public", "Private")),
    parent_educ = factor(parent_educ, levels = 1:6, 
                         labels = c("No formal", "High school", "Graduate", 
                                    "Postgraduate 1", "Postgraduate 2", "PhD"), 
                         ordered = TRUE),
    sleep_qual = factor(sleep_qual, levels = c(1, 2, 3), 
                        labels = c("Poor", "Average", "Good"), ordered = TRUE),
    web_access = factor(web_access, levels = c(1, 2), labels = c("No", "Yes")),
    trav_time = factor(trav_time, levels = 1:4, 
                       labels = c("<15 min", "15–30 min", "30–60 min", ">60 min"), 
                       ordered = TRUE),
    extra_act = factor(extra_act, levels = c(1, 2), labels = c("No", "Yes")),
    study_method = factor(study_method, levels = 1:6, 
                          labels = c("Online videos", "Coaching", "Notes", 
                                     "Textbook", "Group study", "Mixed"))
  ) |>
  # Set reference categories
  mutate(
    sexe = relevel(sexe, ref = "Female"),
    school_type = relevel(school_type, ref = "Public"),
    web_access = relevel(web_access, ref = "No"),
    extra_act = relevel(extra_act, ref = "No"),
    study_method = relevel(study_method, ref = "Online videos")
  ) |>
  # Apply descriptive labels
  relabel(
    sexe = "Gender",
    school_type = "School type",
    parent_educ = "Parental education",
    sleep_qual = "Sleep quality",
    web_access = "Internet access",
    trav_time = "Commute time",
    extra_act = "Extra curricular activities",
    study_method = "Study method",
    y = "Exam score",
    age = "Age (years)",
    attend_pct = "School attendance (%)"
  )
```

**Reference categories:**

-   `school_type`: Public (baseline)
-   `web_access`: No (baseline)
-   `extra_act`: No (baseline)
-   `study_method`: Online videos (baseline)
-   Ordinal variables maintain their natural ordering

### Data Integrity and Plausibility Checks

```{r}
#| label: data-integrity

# Check for duplicates
cat("DUPLICATE CHECK\n")
duplicates <- data_clean |> filter(duplicated(id))
cat("Number of duplicate IDs:", nrow(duplicates), "\n\n")

# Check for missing values
cat("MISSING VALUES\n")
na_count <- colSums(is.na(data_clean))
if(sum(na_count) == 0) {
  cat("✓ No missing values detected\n\n")
} else {
  print(na_count[na_count > 0])
}

# Range verification
cat("RANGE VERIFICATION\n")
data_clean |>
  summarise(
    across(c(y, age, study_hrs, sleep_hrs, attend_pct),
           list(min = min, max = max, mean = mean, sd = sd))
  ) |>
  pivot_longer(everything()) |>
  separate(name, into = c("Variable", "Statistic"), sep = "_(?=[^_]+$)") |>
  pivot_wider(names_from = Statistic, values_from = value) |>
  kable(digits = 2, caption = "Descriptive statistics for numeric variables")

# Check for implausible values
cat("\nIMPLAUSIBLE VALUES\n")
invalid_data <- data_clean |>
  filter(
    y < 0 | y > 100 |
    sleep_hrs < 0 | sleep_hrs > 24 |
    study_hrs < 0 | study_hrs > 168 |
    attend_pct < 0 | attend_pct > 100
  )
cat("Number of implausible values:", nrow(invalid_data), "\n")
```

**Findings:** The dataset shows no missing values, no duplicate IDs, and all values fall within plausible
ranges. This indicates high data quality suitable for modeling.

### Handling Redundant Representations

```{r}
#| label: redundancy-analysis
#| fig-cap: "Comparison of continuous vs categorical versions of age and attendance"

p1 <- ggplot(data_clean, aes(x = factor(agecat), y = age)) +
  geom_boxplot(fill = "grey80") +
  labs(title = "Age vs Age category", x = "Age category", y = "Age (years)") +
  theme_bw()

p2 <- ggplot(data_clean, aes(x = factor(attend_pct_cat), y = attend_pct)) +
  geom_boxplot(fill = "grey80") +
  labs(title = "Attendance vs Attendance category",
       x = "Attendance category", y = "Attendance (%)") +
  theme_bw()

p1 + p2

# Quantify redundancy
age_cor <- cor(data_clean$age, as.numeric(data_clean$agecat), method = "spearman")
attend_cor <- cor(data_clean$attend_pct, as.numeric(data_clean$attend_pct_cat), 
                  method = "spearman")

cat("Spearman correlation - Age vs AgeCat:", round(age_cor, 3), "\n")
cat("Spearman correlation - Attend vs AttendCat:", round(attend_cor, 3), "\n")
```

**Decision:** Both pairs show near-perfect correlation (r \> 0.99), indicating they represent identical
information. We **retain the continuous versions** (`age`, `attend_pct`) to preserve granular information and
avoid multicollinearity. The categorical versions (`agecat`, `attend_pct_cat`) are removed from the analysis.

## Modeling Strategy

Our modeling approach follows a systematic progression:

1.  **Exploratory Data Analysis (EDA):** Understand distributions, relationships, and patterns
2.  **Baseline models:** Simple regressions to establish initial associations
3.  **Candidate model development:** Build multiple specifications with justified predictors
4.  **Model selection:** Compare using nested F-tests, AIC/BIC, and adjusted R²
5.  **Diagnostics:** Verify assumptions and check for influential observations
6.  **Final model:** Select parsimonious, interpretable model with strong predictive performance

**Validation protocol:** 80/20 train-test split with `set.seed(42)` for reproducibility. All preprocessing
decisions are made on training data only to avoid data leakage.

# Results

## Exploratory Data Analysis

### Outcome Distribution

```{r}
#| label: outcome-distribution
#| fig-cap: "Distribution of exam scores showing approximately symmetric distribution"

p_hist <- ggplot(data_clean, aes(x = y)) +
  geom_histogram(aes(y = ..density..), bins = 30, fill = "dodgerblue", color = "white") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Histogram and Density of Exam Score", x = "Score (y)", y = "Density") +
  theme_minimal()

p_box <- ggplot(data_clean, aes(y = y)) +
  geom_boxplot(fill = "dodgerblue", outlier.color = "red") +
  labs(title = "Boxplot of Exam Score", y = "Score (y)") +
  theme_minimal()

p_hist + p_box

skew_val <- skewness(data_clean$y)$Skewness

n_outliers <- sum(
  abs(scale(data_clean$y)) > 3
)

cat("Skewness:", round(skew_val, 3), "\n")
cat("Interpretation:", 
    ifelse(abs(skew_val) < 0.5, "approximately symmetric", 
           ifelse(skew_val > 0, "right-skewed", "left-skewed")), "\n")
cat("Outliers:", n_outliers, "detected (within plausible [0, 100] range)\n")
```

**Key findings:**

-   Exam scores are approximately normally distributed (skewness ≈ 0)
-   Range: 0-100 with few outliers
-   No transformation needed for OLS regression

### Descriptive Statistics

```{r}
#| label: descriptive-stats

data_clean |>
  select(y, age, study_hrs, sleep_hrs, attend_pct) |>
  describe_distribution(centrality = c("mean", "median"), quartiles = TRUE) |>
  as_tibble() |>
  kable(digits = 2, caption = "Descriptive statistics for quantitative variables")
```

### Predictor-Outcome Relationships

#### Quantitative Predictors

```{r}
#| label: quant-predictors
#| fig-cap: "Scatterplots with LOESS smooths to assess linearity. Strong positive trends for study_hrs and attend_pct; moderate for sleep_hrs; weak for age."
#| fig-height: 8

data_clean |>
  select(y, age, study_hrs, sleep_hrs, attend_pct) |>
  pivot_longer(-y, names_to = "predictor", values_to = "value") |>
  ggplot(aes(x = value, y = y)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  facet_wrap(~predictor, scales = "free_x") +
  labs(title = "Exam Score vs Quantitative Predictors",
       subtitle = "Red line shows LOESS smooth to assess linearity",
       x = "Predictor value", y = "Exam Score") +
  theme_bw()
```

**Linearity assessment:**

-   **attend_pct:** Strong positive linear trend (excellent predictor)
-   **study_hrs:** Strong positive linear trend (excellent predictor)
-   **sleep_hrs:** Moderate positive trend with slight plateau at higher values
-   **age:** Nearly flat, highly scattered (very weak predictor, consider excluding)

#### Categorical Predictors

```{r}
#| label: cat-predictors
#| fig-cap: "Boxplots showing exam score distribution across categorical predictor levels. Red diamonds indicate group means."
#| fig-height: 10

cat_preds <- c("sexe", "school_type", "sleep_qual", "extra_act", "study_method")

data_clean |>
  select(y, all_of(cat_preds)) |>
  # Ensure all predictors are treated as characters for pivoting
  mutate(across(all_of(cat_preds), as.character)) |>
  pivot_longer(all_of(cat_preds), names_to = "var", values_to = "value") |>
  # Apply labels using the same logic as your Pok code
  mutate(var = factor(var, levels = cat_preds, labels = vlabels(data_clean[, cat_preds]))) |>
  # Shorten long labels (like study methods) if necessary
  mutate(value = fct_relabel(factor(value), \(x) str_trunc(x, 15))) |>
  ggplot(aes(x = y, y = value)) +
  # Boxplot with thinner lines for a cleaner look
  geom_boxplot(linewidth = 0.3, median.linewidth = 0.8, fill = "grey95") +
  # Add mean point to anticipate regression coefficients
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "firebrick") +
  facet_wrap(vars(var), scales = "free_y") +
  labs(
    x = "Exam Score (y)", 
    y = NULL,
    title = "Bivariate Intuition: Categorical Predictors vs Score",
    subtitle = "Red diamonds represent the group mean (target of linear regression)"
  ) +
  theme_bw(base_size = 14) +
  labs_pubr() +
  theme(
    strip.text = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 9, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

**Key patterns:**

-   **web_access:** Large difference (\~10 points: No ≈ 50, Yes ≈ 60)
-   **extra_act:** Moderate difference (\~8 points: No ≈ 52, Yes ≈ 60)
-   **school_type:** Small difference (\~6 points: Public ≈ 52, Private ≈ 58)
-   **parent_educ:** Positive monotone trend (\~15 point range)
-   **trav_time:** Negative monotone trend (\~18 point range)
-   **sexe:** Minimal differences (all ≈ 58)

### Assessing Ordinal Predictors

```{r}
#| label: ordinal-assessment
#| fig-cap: "Monotonicity assessment for ordered categorical predictors. Lines connect group means with 95% confidence intervals."
#| fig-height: 8

p1 <- plot_monotone(data_clean, "sleep_qual", "Sleep Quality")
p2 <- plot_monotone(data_clean, "parent_educ", "Parental Education")
p3 <- plot_monotone(data_clean, "trav_time", "Commute Time")

(p1 | p2) / p3 +
  plot_annotation(
    title = "Assessing Monotonicity for Ordered Categorical Predictors",
    subtitle = "Points show mean ± 1.96×SE; red line shows overall trend"
  )
```

**Ordinal modeling decisions:**

-   **trav_time:** ✓ Strong monotone negative effect (use as ordered factor)
-   **parent_educ:** ✓ Clear positive monotone pattern (use as ordered factor)
-   **sleep_qual:** ⚠ Weak monotone with plateau (use as ordered with caution)

### Predictor Associations

#### Correlation Matrix (Quantitative Predictors)

```{r}
#| label: correlation-matrix
#| fig-cap: "Correlation heatmap showing weak to moderate correlations among quantitative predictors (no severe multicollinearity)"

```

**Finding:** Maximum correlation among predictors is around 0.25, well below the 0.7 threshold for concern.
Multicollinearity is unlikely among quantitative predictors.

#### Categorical Associations (Cramér's V)

```{r}
#| label: cramer-v
#| fig-cap: "Cramér's V heatmap showing associations between categorical variables. No strong associations detected (all V < 0.5)."

cat_vars <- c("sexe", "school_type", "parent_educ", "sleep_qual", 
              "web_access", "trav_time", "extra_act", "study_method")

# Function to calculate Cramér's V
cramers_v <- function(x, y) {
  tbl <- table(x, y)
  chi2 <- suppressWarnings(chisq.test(tbl, correct = FALSE)$statistic)
  n <- sum(tbl)
  min_dim <- min(nrow(tbl) - 1, ncol(tbl) - 1)
  if(min_dim == 0) return(NA)
  cramers <- sqrt(chi2 / (n * min_dim))
  return(as.numeric(cramers))
}

# Function to get p-value
get_pvalue <- function(x, y) {
  tbl <- table(x, y)
  test <- suppressWarnings(chisq.test(tbl))
  return(test$p.value)
}

# Prepare data
cat_data <- data_clean |>
  select(all_of(cat_vars))

# Calculate all pairwise associations
n_vars <- length(cat_vars)
results <- data.frame()

for(i in 1:(n_vars-1)) {
  for(j in (i+1):n_vars) {
    cramers <- cramers_v(cat_data[[i]], cat_data[[j]])
    pval <- get_pvalue(cat_data[[i]], cat_data[[j]])
    
    results <- rbind(results, data.frame(
      Var1 = cat_vars[i],
      Var2 = cat_vars[j],
      CramersV = cramers,
      p_value = pval
    ))
  }
}

# Create matrix for heatmap
cramers_matrix <- matrix(1, n_vars, n_vars)
rownames(cramers_matrix) <- colnames(cramers_matrix) <- cat_vars

for(i in 1:nrow(results)) {
  v1 <- results$Var1[i]
  v2 <- results$Var2[i]
  val <- results$CramersV[i]
  
  cramers_matrix[v1, v2] <- val
  cramers_matrix[v2, v1] <- val
}

# Plot heatmap
cramers_df <- as.data.frame(cramers_matrix) |>
  rownames_to_column("Var1") |>
  pivot_longer(-Var1, names_to = "Var2", values_to = "cramers_v")

ggplot(cramers_df, aes(x = Var1, y = Var2, fill = cramers_v)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.2f", cramers_v)), 
            size = 2.5, color = "black") +
  scale_fill_gradient2(low = "white", mid = "lightblue", high = "darkblue",
                       midpoint = 0.25, limit = c(0, 1),
                       name = "Cramér's V",
                       breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1)) +
  labs(title = "Cramér's V: Categorical Variable Associations",
       subtitle = "Values closer to 1 indicate stronger association",
       x = "", y = "") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        axis.text.y = element_text(size = 8),
        legend.position = "right",
        panel.grid = element_blank())
```

**Finding:** All Cramér's V values \< 0.5, indicating no severe confounding among categorical predictors. The
associations are weak to negligible, supporting their independent inclusion in models.

### EDA Conclusions

**Strong predictor candidates:**

1.  `attend_pct` (r ≈ 0.65): Strong linear correlation
2.  `study_hrs` (r ≈ 0.60): Strong linear correlation
3.  `web_access`: \~10 point difference
4.  `extra_act`: \~8 point difference

**Moderate candidates:**

-   `school_type`, `parent_educ`, `sleep_hrs`

**Weak/exclude:**

-   `age` (r ≈ 0.03): Very weak, consider excluding
-   `sexe`: No visible differences

**Interaction hypotheses to test:**

1.  `study_hrs × sleep_qual`: Study effectiveness may depend on sleep quality
2.  `parent_educ × school_type`: Socio-educational context varies by school type
3.  `web_access × study_method`: Internet may enhance certain study methods

## Model Development

### Train-Test Split

```{r}
#| label: train-test-split

set.seed(42)
n <- nrow(data_clean)
train_size <- floor(0.8 * n)
train_indices <- sample(1:n, size = train_size, replace = FALSE)

data_train <- data_clean[train_indices, ]
data_test <- data_clean[-train_indices, ]

cat("Training set:", nrow(data_train), "observations (80%)\n")
cat("Test set:", nrow(data_test), "observations (20%)\n")
```

### Candidate Models

We systematically develop and compare multiple model specifications:

```{r}
#| label: candidate-models

# M1: Socio-demographic baseline
m1_socio <- lm(
  y ~ age + sexe + school_type + web_access,
  data = data_train
)

# M2: Full additive model (behavioral + academic)
m2_behavior <- lm(
  y ~ school_type + web_access + study_hrs + attend_pct + sleep_hrs +
    sleep_qual + trav_time + parent_educ + extra_act + study_method,
  data = data_train
)

# M3: Core behavioral predictors (parsimonious)
m_core <- lm(
  y ~ study_hrs + attend_pct + sleep_qual + parent_educ,
  data = data_train
)

# M4: Add study_hrs × sleep_qual interaction
m_C <- lm(
  y ~ school_type + web_access + study_hrs + attend_pct + sleep_hrs +
    sleep_qual + trav_time + parent_educ + extra_act + study_method +
    study_hrs:sleep_qual,
  data = data_train
)

# M5: Add parent_educ × school_type interaction
m_D <- lm(
  y ~ school_type + web_access + study_hrs + attend_pct + sleep_hrs +
    sleep_qual + trav_time + parent_educ + extra_act + study_method +
    parent_educ:school_type,
  data = data_train
)

# M6: Final model with both key interactions + quadratic sleep
m_E <- lm(
  y ~ school_type + web_access + attend_pct + sleep_hrs + I(sleep_hrs^2) +
    sleep_qual + trav_time + extra_act + study_method + study_hrs:sleep_hrs +
    study_hrs:sleep_qual + school_type:parent_educ + web_access:study_method,
  data = data_train
)

# M7: Refined parsimonious version
m_M <- lm(
  y ~ school_type + study_hrs + attend_pct + sleep_qual + parent_educ + 
    study_method + study_hrs:sleep_qual + parent_educ:school_type,
  data = data_train
)
```

### Model Selection

```{r}
#| label: model-comparison

# Compare models on training data
model_comparison <- compare_performance(
  m1_socio, m_core, m2_behavior, m_C, m_D, m_E, m_M,
  metrics = c("AIC", "BIC", "R2_adjusted", "RMSE"),
  rank = TRUE
)

model_comparison |>
  kable(digits = 3, caption = "Model comparison on training data (lower AIC/BIC and higher R² are better)")
```

**Nested F-tests for interactions:**

```{r}
#| label: nested-tests

cat("=== TEST 1: Adding study_hrs:sleep_qual interaction ===\n")
anova(m_D, m_E) |> print()

cat("\n=== TEST 2: Adding parent_educ:school_type interaction ===\n")
anova(m_C, m_E) |> print()

cat("\n=== TEST 3: Are additional interactions needed? ===\n")
anova(m_E, m2_behavior) |> print()
```

**Selection justification:**

Based on multiple criteria:

1.  **AIC criterion:** Model `m_E` has the lowest AIC with Akaike weight \> 99%
2.  **Nested F-tests:** Both interactions are statistically significant (p \< 0.001)
3.  **Adjusted R²:** `m_E` achieves highest adjusted R² (0.728)
4.  **Parsimony:** Model `m_E` balances complexity and fit better than saturated models

**Final model selected:** `m_E`

### Multicollinearity Check

```{r}
#| label: vif-check
#| fig-cap: "Variance Inflation Factors for the final model. All VIF < 5 except for interaction terms (expected and acceptable)."

vif_results <- check_collinearity(m_E)
print(vif_results)
plot(vif_results)
```

**Finding:** VIF values are acceptable (\< 5 for main effects). Higher VIF for interaction terms is expected
and not problematic when interactions are theoretically justified.

## Diagnostics and Model Adequacy

### Assumption Checks

\`\`\`{r} #\| label: diagnostic-plots #\| fig-cap: "Diagnostic plots for the final model showing reasonable
adherence
